---
title: "PHB228 Homework 1"
author: "A69043689"
date: "2026-05-03"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(bench)
library(palmerpenguins)
library(purrr)
```

# Part 1: Vectorization and Data Structures

## 1.1 Vectorization

### a)

```{r}
# TODO: Define vectorized_zscore().

# TODO: Define loop_zscore().

# TODO: Test both functions on a numeric column from iris.
```

### b)

```{r}
# TODO: Use bench::mark() with at least 50 iterations.

# TODO: Create a bar plot comparing median execution times.
```

### c)

```{r}
# TODO: Use ChickWeight to create Growth.Rate and Weight.Gain.Percent.

# TODO: Summarize average growth rate and mean maximum weight by Diet.
```

## 1.2 Data Structures and Functional Programming

### a)

```{r}
# TODO: Split palmerpenguins::penguins into a list by species.

# TODO: Add a sample size attribute to each species data frame.
```

### b)

```{r}
# TODO: Use lapply() to calculate numeric means excluding NAs.

# TODO: Redo a comparable calculation with purrr::map_dbl().
```

### c)

```{r}
# TODO: Use tapply() for mean body mass by species.

# TODO: Use tapply() for mean body mass by species and sex.
```

### d)

```{r}
# TODO: Demonstrate copy-on-modify semantics.
```

Write 2-3 sentences explaining what happened.

### e)

```{r}
# TODO: Time the inefficient growing-vector loop.

# TODO: Rewrite using preallocation and compare with system.time().
```

Write 2-3 sentences explaining why preallocation is faster.

# Part 2: Matrix Operations and Decompositions

## 2.1 Matrix Operations

```{r}
A <- matrix(c(4, 2, 1, 5), nrow = 2)
B <- matrix(c(3, 1, 0, 2), nrow = 2)
C <- matrix(c(1, 3, 5, 2, 4, 6), nrow = 2)
```

### a)

```{r}
# TODO: Calculate A + B.

# TODO: Calculate A * B.

# TODO: Calculate A %*% B.

# TODO: Calculate t(A) %*% C.

# TODO: Calculate solve(A).
```

Briefly explain each operation.

### b)

```{r}
# TODO: Create is_symmetric().

# TODO: Test it on A, B, and t(A) %*% A.
```

## 2.2 Eigenvalue Decomposition

```{r}
M <- matrix(c(4, 2, 2, 3), nrow = 2)
```

### a)

```{r}
# TODO: Compute eigenvalues and eigenvectors with eigen().
```

### b)

```{r}
# TODO: Verify Mx = lambda x for each eigenpair.
```

### c)

```{r}
# TODO: Reconstruct M using P %*% D %*% solve(P).

# TODO: Verify the reconstruction matches M.
```

## 2.3 SVD and QR Decomposition

### a) SVD

```{r}
X <- matrix(c(3, 2, 2, 2, 3, -2, 2, -2, 3), nrow = 3)

# TODO: Compute SVD.

# TODO: Create a rank-1 approximation.

# TODO: Calculate the Frobenius norm of the difference.
```

Explain what information is lost in the rank-1 approximation.

### b) QR and Least Squares

```{r}
set.seed(228)
n <- 100
x <- runif(n, -5, 5)
y <- 3 + 2 * x + rnorm(n, sd = 1)

# TODO: Create the design matrix.

# TODO: Solve least squares via QR decomposition.

# TODO: Compare with lm().
```

Explain the advantages of QR decomposition for least squares.

# Part 3: Optimization and Simulation

## 3.1 Gradient Descent

### a)

```{r}
# TODO: Write rosenbrock(x).
```

### b)

```{r}
# TODO: Write rosenbrock_grad(x).
```

### c)

```{r}
# TODO: Implement gradient_descent().
```

### d)

```{r}
# TODO: Apply gradient_descent() from c(-1.2, 1.0).
```

Report the solution, function value, and iterations.

## 3.2 Comparison with Built-in Methods

### a)

```{r}
# TODO: Use optim() with method = "BFGS".
```

### b)

```{r}
# TODO: Use optim() with method = "Nelder-Mead".

# TODO: Compare gradient descent, BFGS, and Nelder-Mead.
```

Discuss solution accuracy, iterations, and function evaluations.

## 3.3 Simulation Study

### a)

```{r}
# TODO: Write run_simulation(n_samples, n_reps, dist_type).
```

### b)

```{r}
# TODO: Run simulations for normal, t(df = 3), and contaminated normal.

# TODO: Calculate bias, variance, and MSE for each estimator.
```

### c)

```{r}
# TODO: Create a summary table.
```

Write 3-5 sentences interpreting estimator performance.

# Part 4: Bootstrap Methods

## 4.1 Bootstrap Confidence Intervals

### a)

```{r}
set.seed(42)
returns <- rnorm(50, mean = 0.08, sd = 0.12)
returns <- returns + rexp(50, rate = 20) - 0.05

# TODO: Implement bootstrap_mean().

# TODO: Report percentile, basic, and normal intervals.
```

### b)

```{r}
# TODO: Plot the bootstrap distribution and add vertical CI lines.
```

Discuss which CI method is most appropriate and why.

## 4.2 Bootstrap Hypothesis Testing

### a)

```{r}
set.seed(123)
control <- rnorm(40, mean = 10, sd = 2)
treatment <- rnorm(40, mean = 11.5, sd = 2)

# TODO: Implement bootstrap_ttest().

# TODO: Plot the bootstrap null distribution with observed difference.
```

### b)

```{r}
# TODO: Compare bootstrap test results with t.test().
```

Discuss differences and potential advantages of the bootstrap approach.

# Academic Integrity

I used AI assistance to organize the repository and create a Quarto response template. All final code, analysis, and interpretations should be completed and checked individually before submission.
