---
title: "PHB228 Homework 1"
author: "A69043689"
date: "2026-05-03"
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(bench)
library(palmerpenguins)
library(purrr)
```

# Part 1: Vectorization and Data Structures

## 1.1 Vectorization

### a)

```{r}
vectorized_zscore <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

loop_zscore <- function(x) {
  x_mean <- mean(x, na.rm = TRUE)
  x_sd <- sd(x, na.rm = TRUE)
  z <- numeric(length(x))

  for (i in seq_along(x)) {
    z[i] <- (x[i] - x_mean) / x_sd
  }

  z
}

iris_x <- iris$Sepal.Length
head(tibble(
  original = iris_x,
  vectorized = vectorized_zscore(iris_x),
  loop = loop_zscore(iris_x)
))

all.equal(vectorized_zscore(iris_x), loop_zscore(iris_x))
```

### b)

```{r}
zscore_bench <- bench::mark(
  vectorized = vectorized_zscore(iris_x),
  loop = loop_zscore(iris_x),
  iterations = 50,
  check = TRUE
)

zscore_bench |> select(expression, median, `itr/sec`, mem_alloc)

zscore_bench |>
  mutate(
    implementation = as.character(expression),
    median_seconds = as.numeric(median)
  ) |>
  ggplot(aes(x = implementation, y = median_seconds, fill = implementation)) +
  geom_col(show.legend = FALSE) +
  labs(
    title = "Median execution time for z-score functions",
    x = "Implementation",
    y = "Median time (seconds)"
  )
```

### c)

```{r}
chick_growth <- ChickWeight |>
  arrange(Chick, Time) |>
  group_by(Chick) |>
  mutate(
    first_weight = first(weight),
    first_time = first(Time),
    last_weight = last(weight),
    Growth.Rate = if_else(
      Time == first_time,
      NA_real_,
      (weight - first_weight) / (Time - first_time)
    ),
    Weight.Gain.Percent = (last_weight - first_weight) / first_weight * 100
  ) |>
  ungroup()

chick_growth |>
  group_by(Diet) |>
  summarize(
    average_growth_rate = mean(Growth.Rate, na.rm = TRUE),
    mean_maximum_weight = mean(tapply(weight, Chick, max)[as.character(unique(Chick))]),
    mean_weight_gain_percent = mean(Weight.Gain.Percent),
    .groups = "drop"
  )
```

## 1.2 Data Structures and Functional Programming

### a)

```{r}
penguin_by_species <- split(penguins, penguins$species)

penguin_by_species <- lapply(penguin_by_species, function(df) {
  attr(df, "sample_size") <- nrow(df)
  df
})

map_int(penguin_by_species, ~ attr(.x, "sample_size"))
```

### b)

```{r}
numeric_means_lapply <- lapply(penguin_by_species, function(df) {
  numeric_cols <- df |> select(where(is.numeric))
  lapply(numeric_cols, mean, na.rm = TRUE)
})

numeric_means_lapply

body_mass_means_map <- penguin_by_species |>
  map_dbl(~ mean(.x$body_mass_g, na.rm = TRUE))

body_mass_means_map
```

### c)

```{r}
tapply(penguins$body_mass_g, penguins$species, mean, na.rm = TRUE)

tapply(
  penguins$body_mass_g,
  list(species = penguins$species, sex = penguins$sex),
  mean,
  na.rm = TRUE
)
```

### d)

```{r}
x <- c(1, 2, 3, 4)
y <- x

before <- tibble(object = c("x", "y"), value = c(paste(x, collapse = ", "), paste(y, collapse = ", ")))
y[1] <- 99
after <- tibble(object = c("x", "y"), value = c(paste(x, collapse = ", "), paste(y, collapse = ", ")))

list(before = before, after = after)
```

After assigning `y <- x`, both names initially point to the same vector value. When `y[1]` is changed, R creates a modified copy for `y`, so `x` keeps its original values. This copy-on-modify behavior makes ordinary assignment behave safely while avoiding unnecessary copies until a change is made.

### e)

```{r}
growing_time <- system.time({
  result_growing <- numeric(0)
  for (i in 1:10000) {
    result_growing <- c(result_growing, i^2)
  }
})

preallocated_time <- system.time({
  result_preallocated <- numeric(10000)
  for (i in 1:10000) {
    result_preallocated[i] <- i^2
  }
})

rbind(growing_vector = growing_time, preallocated = preallocated_time)
all.equal(result_growing, result_preallocated)
```

The growing-vector version repeatedly allocates a new vector and copies all previous values each time `c()` is called. The preallocated version reserves the needed memory once and fills positions in place, so it avoids thousands of repeated allocations and copies. That is why the preallocated loop is faster and scales better.

# Part 2: Matrix Operations and Decompositions

## 2.1 Matrix Operations

```{r}
A <- matrix(c(4, 2, 1, 5), nrow = 2)
B <- matrix(c(3, 1, 0, 2), nrow = 2)
C <- matrix(c(1, 3, 5, 2, 4, 6), nrow = 2)
```

### a)

```{r}
A_plus_B <- A + B
A_elementwise_B <- A * B
A_matrix_B <- A %*% B
tA_matrix_C <- t(A) %*% C
A_inverse <- solve(A)

list(
  A_plus_B = A_plus_B,
  A_elementwise_B = A_elementwise_B,
  A_matrix_B = A_matrix_B,
  tA_matrix_C = tA_matrix_C,
  A_inverse = A_inverse
)
```

`A + B` adds corresponding entries of the two matrices. `A * B` multiplies corresponding entries element by element, while `A %*% B` performs matrix multiplication using row-by-column inner products. `t(A) %*% C` first transposes `A` and then matrix-multiplies by `C`; `solve(A)` returns the inverse matrix of `A`.

### b)

```{r}
is_symmetric <- function(mat, tol = .Machine$double.eps^0.5) {
  if (!is.matrix(mat)) {
    stop("Input must be a matrix.")
  }
  if (nrow(mat) != ncol(mat)) {
    stop("Input matrix must be square.")
  }
  isTRUE(all.equal(mat, t(mat), tolerance = tol))
}

tibble(
  matrix = c("A", "B", "t(A) %*% A"),
  symmetric = c(is_symmetric(A), is_symmetric(B), is_symmetric(t(A) %*% A))
)
```

## 2.2 Eigenvalue Decomposition

```{r}
M <- matrix(c(4, 2, 2, 3), nrow = 2)
```

### a)

```{r}
M_eigen <- eigen(M)
M_eigen
```

### b)

```{r}
eigen_checks <- map(seq_along(M_eigen$values), function(i) {
  lambda <- M_eigen$values[i]
  x_vec <- M_eigen$vectors[, i]

  tibble(
    eigenpair = i,
    left_side = list(M %*% x_vec),
    right_side = list(lambda * x_vec),
    matches = isTRUE(all.equal(as.vector(M %*% x_vec), as.vector(lambda * x_vec)))
  )
})

bind_rows(eigen_checks)
```

### c)

```{r}
P <- M_eigen$vectors
D <- diag(M_eigen$values)
M_reconstructed <- P %*% D %*% solve(P)

M_reconstructed
all.equal(M, M_reconstructed)
```

## 2.3 SVD and QR Decomposition

### a) SVD

```{r}
X <- matrix(c(3, 2, 2, 2, 3, -2, 2, -2, 3), nrow = 3)

X_svd <- svd(X)
X_svd

X_rank1 <- X_svd$d[1] * X_svd$u[, 1, drop = FALSE] %*% t(X_svd$v[, 1, drop = FALSE])
frobenius_norm <- norm(X - X_rank1, type = "F")

X_rank1
frobenius_norm
```

The rank-1 approximation keeps only the strongest singular direction, so it preserves the largest linear pattern in `X`. The remaining singular directions are discarded, which means smaller-scale variation and structure orthogonal to the dominant direction are lost. The Frobenius norm measures the total size of that discarded information.

### b) QR and Least Squares

```{r}
set.seed(228)
n <- 100
x <- runif(n, -5, 5)
y <- 3 + 2 * x + rnorm(n, sd = 1)

design_matrix <- cbind(intercept = 1, x = x)
qr_fit <- qr(design_matrix)
beta_qr <- qr.coef(qr_fit, y)
beta_lm <- coef(lm(y ~ x))

tibble(
  method = c("QR", "lm"),
  intercept = c(beta_qr[1], beta_lm[1]),
  slope = c(beta_qr[2], beta_lm[2])
)
```

QR decomposition solves least squares without explicitly forming `(X'X)^{-1}`. This is numerically more stable, especially when predictors are correlated or the design matrix is close to singular. It also provides a direct way to assess matrix rank and solve overdetermined systems efficiently.

# Part 3: Optimization and Simulation

## 3.1 Gradient Descent

### a)

```{r}
rosenbrock <- function(x) {
  (1 - x[1])^2 + 100 * (x[2] - x[1]^2)^2
}
```

### b)

```{r}
rosenbrock_grad <- function(x) {
  c(
    -2 * (1 - x[1]) - 400 * x[1] * (x[2] - x[1]^2),
    200 * (x[2] - x[1]^2)
  )
}
```

### c)

```{r}
gradient_descent <- function(init, grad_fn, learning_rate = 0.001,
                             max_iter = 10000, tol = 1e-6,
                             fn = rosenbrock) {
  x_current <- init

  for (iter in seq_len(max_iter)) {
    grad <- grad_fn(x_current)
    grad_norm <- sqrt(sum(grad^2))

    if (grad_norm < tol) {
      return(list(solution = x_current, value = fn(x_current), iterations = iter - 1))
    }

    x_current <- x_current - learning_rate * grad
  }

  list(solution = x_current, value = fn(x_current), iterations = max_iter)
}
```

### d)

```{r}
gd_result <- gradient_descent(
  init = c(-1.2, 1.0),
  grad_fn = rosenbrock_grad,
  learning_rate = 0.001,
  max_iter = 10000,
  tol = 1e-6
)

gd_result
```

The gradient descent solution is `r paste(round(gd_result$solution, 6), collapse = ", ")`, with function value `r signif(gd_result$value, 6)` after `r gd_result$iterations` iterations.

## 3.2 Comparison with Built-in Methods

### a)

```{r}
bfgs_result <- optim(
  par = c(-1.2, 1.0),
  fn = rosenbrock,
  gr = rosenbrock_grad,
  method = "BFGS"
)

bfgs_result
```

### b)

```{r}
nelder_mead_result <- optim(
  par = c(-1.2, 1.0),
  fn = rosenbrock,
  method = "Nelder-Mead"
)

optimization_comparison <- tibble(
  method = c("Gradient descent", "BFGS", "Nelder-Mead"),
  x1 = c(gd_result$solution[1], bfgs_result$par[1], nelder_mead_result$par[1]),
  x2 = c(gd_result$solution[2], bfgs_result$par[2], nelder_mead_result$par[2]),
  function_value = c(gd_result$value, bfgs_result$value, nelder_mead_result$value),
  iterations = c(gd_result$iterations, bfgs_result$counts["gradient"], NA),
  function_evaluations = c(NA, bfgs_result$counts["function"], nelder_mead_result$counts["function"])
)

optimization_comparison
```

BFGS reaches the known minimum near `(1, 1)` with a very small function value using far fewer evaluations because it uses gradient and curvature information. Nelder-Mead also gets close to the optimum but usually needs more function evaluations because it does not use derivatives. The fixed-step gradient descent implementation improves steadily but is much slower on the Rosenbrock function's narrow curved valley, so after 10,000 iterations it is less accurate than the built-in optimizers.

## 3.3 Simulation Study

### a)

```{r}
run_simulation <- function(n_samples, n_reps, dist_type) {
  map_dfr(seq_len(n_reps), function(rep_id) {
    sample_values <- switch(
      dist_type,
      normal = rnorm(n_samples, mean = 0, sd = 1),
      t3 = rt(n_samples, df = 3),
      contaminated = if_else(
        runif(n_samples) < 0.9,
        rnorm(n_samples, mean = 0, sd = 1),
        rnorm(n_samples, mean = 0, sd = 10)
      ),
      stop("Unknown distribution type.")
    )

    tibble(
      replication = rep_id,
      distribution = dist_type,
      mean = mean(sample_values),
      median = median(sample_values),
      trimmed_mean = mean(sample_values, trim = 0.1)
    )
  })
}
```

### b)

```{r}
set.seed(228)
simulation_results <- bind_rows(
  run_simulation(30, 1000, "normal"),
  run_simulation(30, 1000, "t3"),
  run_simulation(30, 1000, "contaminated")
)

simulation_summary <- simulation_results |>
  pivot_longer(
    cols = c(mean, median, trimmed_mean),
    names_to = "estimator",
    values_to = "estimate"
  ) |>
  group_by(distribution, estimator) |>
  summarize(
    bias = mean(estimate),
    variance = var(estimate),
    mse = mean(estimate^2),
    .groups = "drop"
  )

simulation_summary
```

### c)

```{r}
simulation_summary |>
  arrange(distribution, mse)
```

Under the normal distribution, the sample mean has the lowest or nearly lowest MSE because it uses all observations efficiently when there are no heavy tails or outliers. Under the t distribution with 3 degrees of freedom, the median and trimmed mean are more competitive because they reduce the influence of extreme observations. Under the contaminated normal distribution, the trimmed mean performs especially well because it keeps more information than the median while still downweighting extreme contaminated values. Bias is small for all three estimators because all simulated distributions are centered at zero, so the main performance difference is driven by variance.

# Part 4: Bootstrap Methods

## 4.1 Bootstrap Confidence Intervals

### a)

```{r}
set.seed(42)
returns <- rnorm(50, mean = 0.08, sd = 0.12)
returns <- returns + rexp(50, rate = 20) - 0.05

bootstrap_mean <- function(x, B = 2000, conf = 0.95) {
  theta_hat <- mean(x)
  n <- length(x)
  boot_means <- replicate(B, mean(sample(x, size = n, replace = TRUE)))
  alpha <- 1 - conf
  se <- sd(boot_means)

  percentile <- quantile(boot_means, probs = c(alpha / 2, 1 - alpha / 2), names = FALSE)
  basic <- 2 * theta_hat - rev(percentile)
  normal <- theta_hat + qnorm(c(alpha / 2, 1 - alpha / 2)) * se

  list(
    estimate = theta_hat,
    se = se,
    boot_means = boot_means,
    intervals = tibble(
      method = c("percentile", "basic", "normal"),
      lower = c(percentile[1], basic[1], normal[1]),
      upper = c(percentile[2], basic[2], normal[2])
    )
  )
}

set.seed(228)
boot_returns <- bootstrap_mean(returns, B = 2000)
boot_returns$estimate
boot_returns$se
boot_returns$intervals
```

### b)

```{r}
boot_plot_data <- tibble(boot_mean = boot_returns$boot_means)
ci_lines <- boot_returns$intervals |>
  pivot_longer(cols = c(lower, upper), names_to = "bound", values_to = "value")

ggplot(boot_plot_data, aes(x = boot_mean)) +
  geom_histogram(bins = 35, color = "white", fill = "steelblue") +
  geom_vline(data = ci_lines, aes(xintercept = value, color = method, linetype = bound), linewidth = 0.8) +
  geom_vline(xintercept = boot_returns$estimate, color = "black", linewidth = 0.8) +
  labs(
    title = "Bootstrap distribution of the mean return",
    x = "Bootstrap mean",
    y = "Count",
    color = "CI method",
    linetype = "Bound"
  )
```

The percentile interval is appropriate here because it uses the empirical bootstrap distribution directly and does not require a strong normality assumption. The returns include exponential noise, so the sampling distribution can be slightly asymmetric. The normal interval is easy to interpret but can be less reliable when the bootstrap distribution is skewed. The basic interval is also useful, but its reflection around the observed estimate can be less intuitive for skewed data.

## 4.2 Bootstrap Hypothesis Testing

### a)

```{r}
set.seed(123)
control <- rnorm(40, mean = 10, sd = 2)
treatment <- rnorm(40, mean = 11.5, sd = 2)

bootstrap_ttest <- function(control, treatment, B = 2000) {
  observed_diff <- mean(treatment) - mean(control)
  pooled_mean <- mean(c(control, treatment))

  control_null <- control - mean(control) + pooled_mean
  treatment_null <- treatment - mean(treatment) + pooled_mean

  boot_diffs <- replicate(B, {
    boot_control <- sample(control_null, length(control_null), replace = TRUE)
    boot_treatment <- sample(treatment_null, length(treatment_null), replace = TRUE)
    mean(boot_treatment) - mean(boot_control)
  })

  list(
    observed_diff = observed_diff,
    boot_diffs = boot_diffs,
    p_value = mean(abs(boot_diffs) >= abs(observed_diff))
  )
}

set.seed(228)
boot_test <- bootstrap_ttest(control, treatment, B = 2000)
boot_test$observed_diff
boot_test$p_value

ggplot(tibble(diff = boot_test$boot_diffs), aes(x = diff)) +
  geom_histogram(bins = 35, color = "white", fill = "gray55") +
  geom_vline(xintercept = boot_test$observed_diff, color = "red", linewidth = 1) +
  geom_vline(xintercept = -boot_test$observed_diff, color = "red", linewidth = 1, linetype = "dashed") +
  labs(
    title = "Bootstrap null distribution for difference in means",
    x = "Mean treatment - mean control under null",
    y = "Count"
  )
```

### b)

```{r}
t_test_result <- t.test(treatment, control)

tibble(
  method = c("Bootstrap", "Welch t-test"),
  observed_difference = c(boot_test$observed_diff, diff(rev(t_test_result$estimate))),
  p_value = c(boot_test$p_value, t_test_result$p.value)
)
```

The bootstrap and Welch t-test both indicate a statistically meaningful positive difference between the treatment and control means for these simulated data. Their p-values may differ slightly because the bootstrap uses the empirical resampling distribution rather than the theoretical t approximation. The bootstrap approach is useful when normality or equal-variance assumptions are questionable, and it can be adapted to statistics whose standard errors are harder to derive analytically. The t-test remains efficient and simple when its assumptions are reasonable.

# Academic Integrity

I used AI assistance to organize the repository, complete the Quarto response template, and check rendering. I reviewed the code, output, and written interpretations before submission.
